---
title: "Geographic data analysis in R and Python: comparing code and outputs for vector data"
format: html
# # Alternative title:
# title: "Teaching R and Python for geographic data: shared experiences and lessons learned"
author: 
  - name: Robin Lovelace
    affiliation: University of Leeds, Active Travel England, UK
    orcid: 0000-0001-5679-6536
    email: 
  - name: Anita Graser
    affiliation: Austrian Institute of Technology, Austria
    orcid: 0000-0001-5361-2885
    email: 
  - name: Michael Dorman
    affiliation: Ben-Gurion University of the Negev, Israel
    orcid: 0000-0001-6450-8047
    email:
  - name: Jakub Nowosad
    affiliation: Adam Mickiewicz University, Poznań, Poland
    orcid: 0000-0002-1057-3721
    email: 
---

```{r}
#| include: false
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
# install python packages needed for this document:
pkgs = c("geopandas", "shapely", "pandas", "rasterio", "matplotlib", "folium", "mapclassify")
reticulate::py_install(pkgs)
```

# Introduction

In this blog post, we talk about our experience teaching R and Python for geocomputation.
The focus of the blog post is on geographic vector data, meaning points, lines, polygons (and their 'multi' variants) and the attributes associated with them.
Geographic data analysis is a broad topic and in a later post we will cover raster data, meaning gridded data such as satellite images.

 <!---
 broadly defined as [follows](https://r.geocompx.org/intro#what-is-geocomputation):

> Working with geographic data in a computational way, focusing on code, reproducibility and modularity. 
--->
The context of this blog post is the [OpenGeoHub Summer School 2023](https://opengeohub.org/summer-school/opengeohub-summer-school-poznan-2023/) which has courses on R, Python and Julia.
The size and the diversity of the event has grown over the years.
Noting that many events focus on just one language, and the advantages of diversity of languages and approaches, we wanted to follow-up in a blog post that could be useful to others.

OpenGeoHub 2023 was also a unique chance for the authors of the in-progress open source book, [Geocomputation with Python](https://py.geocompx.org/) to meet in person: the first time we have all been in one place at the same time.

The post is based on the following lecture notes, which we recommend checking out for deeper dives into the R and Python implementations of geocomputation:

- [Tidy geographic data with sf, dplyr, ggplot2, geos and friends](https://ogh23.robinlovelace.net/tidy)
- [Working with spatial data in Python](https://geobgu.xyz/presentations/p_2023_ogh/)

<!---
 TODO: Add video links 
--->

# Comparing R and Python for vector geographic data analysis

## Loading packages

We will start by loading core packages for working with geographic vector and attribute data.
See detailed description of [R](https://ogh23.robinlovelace.net/tidy#vector-data) and [Python](https://geobgu.xyz/presentations/p_2023_ogh/01-vector.html) implementations in the respective lecture note sections.

::: {.panel-tabset group="language"}

## Python

```{python}
import pandas as pd
from shapely import Point
import geopandas as gpd
```

## R

<!---
 jn: I would suggest to use specific packages from tidyverse instead of attaching the whole tidyverse 

rl: Why? Tidyverse is popular and it makes life easy.
Also that's the approach in the teaching materials.
For the book and for software development that's another matter but for this blog post I think it's fine.
--->

```{r}
library(sf)
library(tidyverse)
library(tmap)
```

:::

## Creating geographic data

The following commands create geographic datasets 'from scratch' representing coordinates of a the faculty where the Summer School takes place, and few hotels, in Poznan, Poland.
Most projects start with pre-generated data, but it's useful to create datasets to understand data structures.

::: {.panel-tabset group="language"}

## Python

```{python}
poi = gpd.GeoDataFrame([
    {"name": "Faculty",        "geometry": Point(16.9418, 52.4643)},
    {"name": "Hotel ForZa",    "geometry": Point(16.9474, 52.4436)},
    {"name": "Hotel Lechicka", "geometry": Point(16.9308, 52.4437)},
    {"name": "FairPlayce",     "geometry": Point(16.9497, 52.4604)},
], crs=4326)
```

## R

```{r}
poi_df = tribble(
  ~name, ~lon, ~lat,
  "Faculty",        16.9418, 52.4643,
  "Hotel ForZa",    16.9474, 52.4436,
  "Hotel Lechicka", 16.9308, 52.4437,
  "FairPlayce",     16.9497, 52.4604
)
poi_sf = sf::st_as_sf(poi_df, coords = c("lon", "lat"), crs = "EPSG:4326")
```

:::

### Downloading data

The following commands download data from the internet.

::: {.panel-tabset group="language"}

## Python

```{python}
import urllib.request
import zipfile
import os
u = "https://github.com/Robinlovelace/opengeohub2023/releases/download/data/data.zip"
f = os.path.basename(u)
if not os.path.exists("data"):
    urllib.request.urlretrieve(u, f)
```

## R

```{r}
u = "https://github.com/Robinlovelace/opengeohub2023/releases/download/data/data.zip"
f = basename(u)
if (!dir.exists("data")) {
  download.file(u, f)
}
```

Note that we can read directly from ZIP files with R and Python thanks to the [GDAL Virtual File System](https://gdal.org/user/virtual_file_systems.html).

:::

## Unzipping data

We can unzip the data in R and Python: it contains spatial information about bus stops.

::: {.panel-tabset group="language"}

## Python

```{python}
with zipfile.ZipFile(f, 'r') as zip_ref:
    zip_ref.extractall()
```

## R

```{r}
unzip(f)
```

:::

## Reading and printing geographic data

As shown below, Python and R implemenations to import a shapefile are similar.
Note: we recommend using open file formats such as GeoPackage (`.gpkg`), as outlined in [Geocomputation with R](https://r.geocompx.org/read-write) and [Geocomputation with Python](https://py.geocompx.org/08-read-write-plot).

::: {.panel-tabset group="language"}

## Python

```{python}
pol_all = gpd.read_file("zip://data.zip!data/osm/gis_osm_transport_a_free_1.shp")
pol_all
```

## R

```{r}
pol_all = sf::read_sf("/vsizip/data.zip/data/osm/gis_osm_transport_a_free_1.shp")
pol_all
```

:::

Note: in R, you can read-in the dataset from the URL in a single line of code without first downloading the zip file:

```{r}
#| eval: false
uz = paste0("/vsizip//vsicurl/", u, "/data/osm/gis_osm_transport_a_free_1.shp")
pol_all = sf::read_sf(uz)
```

## Subsetting by attributes

The following commands select a subset of the data based on attribute values (looking for a specific string in the `name` column).

::: {.panel-tabset group="language"}

## Python

```{python}
pol = pol_all[pol_all['name'].str.contains('Port*.+Poz', na=False)]
pol
```

## R

```{r}
pol = pol_all |>
  filter(str_detect(name, "Port*.+Poz"))
pol
```

:::

## Basic plotting

The following commands plot the data.
Note that by default, R's `plot()` method for `{sf}` objects creates a plot for each column in the data (up to 9 by default). <!--- Perhaps switch to plot(st_geometry(pol)) in R, to make it more comparable with Python? --->

::: {.panel-tabset group="language"}

## Python

```{python}
pol.plot();
```

## R

```{r}
plot(pol)
```

:::

The arguments needed to change the colour of the fill and border are different in R and Python, but the results are similar.

::: {.panel-tabset group="language"}

## Python

```{python}
pol.plot(color='none', edgecolor='black');
```

## R

```{r}
plot(st_geometry(pol), col = "white", border = "black")
```

:::

## Creating geographic data frames from a CSV file

The following commands create a geographic data frame from a CSV file.
Note that two steps---creating the geometry column and combining it with the original table, hereby combined into one complex expression---are needed to convert a `DataFrame` to a `GeoDataFrame` in Python, whereas in R the `sf::st_as_sf()` function can be used to convert a `data.frame` to a spatial data frame directly.

::: {.panel-tabset group="language"}

## Python

```{python}
# Unzip the data.zip file:
with zipfile.ZipFile(f, 'r') as zip_ref:
    zip_ref.extractall("data")
stops = pd.read_csv("data/gtfs/stops.txt")
stops = gpd.GeoDataFrame(
    stops.drop(columns=['stop_lon', 'stop_lat', 'stop_code']),
    geometry = gpd.points_from_xy(stops.stop_lon, stops.stop_lat),
    crs = 4326)
stops
```

## R

```{r}
stops = read_csv("data/gtfs/stops.txt") |>
  select(-stop_code) |>
  st_as_sf(coords = c("stop_lon", "stop_lat"), crs = "EPSG:4326")
stops
```

:::

## Plotting attributes and layers

The following commands plot the bus stops loaded in the previous step.
Note that the **tmap** package is hereby used in R to create these more advanced plots, as it also supports interactive mapping (see below).

::: {.panel-tabset group="language"}

## Python

```{python}
stops.plot(markersize=1, column='zone_id', legend=True);
```

## R

```{r}
tm_shape(stops) +
  tm_symbols(size = 0.1, col = "zone_id")
```

:::

We can add basic overlays in both languages as follows.

::: {.panel-tabset group="language"}

## Python

```{python}
base = stops.plot(markersize=0.1)
poi.plot(ax=base, color='red');
```

## R

<!---
 jn: should the next example use tmap instead of base R? 
--->

```{r}
plot(stops$geometry, col = "grey", pch = 20, cex = 0.5)
plot(poi_sf$geometry, col = "red", add = TRUE)
```

:::

## Interactive plots

The following commands create interactive plots, in Python and R respectively.
The Python code requires the **folium** and **mapclassify** packages, which are not installed by default when you install **geopandas**.
Note that with **tmap**, you can use the same code to create static and interactive plots, by changing the `tmap_mode()`. 

::: {.panel-tabset group="language"}

## Python

```{python}
stops.explore(column='zone_id', legend=True, cmap='Dark2')
```

## R

```{r}
tmap_mode("view")
tm_shape(stops) +
  tm_symbols(size = 0.1, col = "zone_id")
```

:::

## Reprojecting data

The following commands reproject the data to a local projected Coordinate Reference System (CRS).

::: {.panel-tabset group="language"}

## Python

```{python}
poi.crs
poi_projected = poi.to_crs(2180)
stops_projected = stops.to_crs(2180)
```

## R

```{r}
st_crs(poi_sf)
poi_projected = st_transform(poi_sf, 2180)
stops_projected = st_transform(stops, 2180)
```

:::

## Buffers

The following commands create buffers around the points.
Note that R allows buffer to be created directly from a spatial data frame with geographic (lon/lot) coordinates thanks to its integration with Google's S2 spherical geometry engine, as outlined in [Geocomputation with R](https://r.geocompx.org/reproj-geo-data#geom-proj).
For buffer operations to work in Python you must reproject the data first (which we did, see above) (although there are plans for **geopandas** to support a spherical geometry backend at some point, as discussed in issue [#2098](https://github.com/geopandas/geopandas/issues/2098)).

::: {.panel-tabset group="language"}

## Python

<!---
 TODO: can we improve this? 
--->

To create a new vector layers named `poi_buffer`, in both languages, we can do the following.

```{python}
poi_buffer = poi.copy()
poi_buffer.geometry = poi_projected.buffer(150).to_crs(4326)
```

## R

```{r}
poi_buffer = st_buffer(poi_sf, 150)
```

:::

## Calculating distances and areas

An interesting difference between R and Python is that the former uses the `units` package to store units, making it easy to convert between them, as outlined in the [buffers section of the R lecture notes](https://ogh23.robinlovelace.net/tidy#buffers). 

::: {.panel-tabset group="language"}

## Python

```{python}
poi_buffer.to_crs(2180).area
```

## R

```{r}
st_area(poi_buffer)
```

:::

## Spatial subsetting

Code to subset the bus stops within the buffered `poi` points is shown below.
The R code is more concise because there is a special `[` notation for the specific case of subsetting by intersection. In Python you must undertake the explicit steps, which are applicable to any predicate in both languages:

- Take the unary union of the buffered points before subsetting
- Create a boolean `Series` object with the `.intersects` or other method, and use the boolean Series to subset the data (rather than another geographic object)

::: {.panel-tabset group="language"}

## Python

```{python}
poi_union = poi_buffer.unary_union
sel = stops.intersects(poi_union)
stops_in_b = stops[sel]
stops_in_b
```

## R

```{r}
stops_in_b = stops[poi_buffer, ]
stops_in_b
```

:::

## Spatial joins

Spatial joins are implemented with similar functions in R and Python and the outputs are the same.
See the [Python](https://geobgu.xyz/presentations/p_2023_ogh/01-vector.html#sec-spatial-join) and R tutorials, and in Geocomputation with R Section [4.2.4](https://r.geocompx.org/spatial-operations#spatial-joining) and Geocomputation with Python [3.3.4](https://py.geocompx.org/04-spatial-operations#sec-spatial-joining) for more details.

::: {.panel-tabset group="language"}

## Python

```{python}
poi_buffer.sjoin(stops, how='left')
```

## R

```{r}
st_join(poi_buffer, stops)
```

:::

```{python}
#| eval: false
#| echo: false
# Aim: check to see if there's a bug in sjoin
points = gpd.GeoDataFrame([
    {"name": "a",        "geometry": Point(0, 0)},
    {"name": "b",    "geometry": Point(1, 0)},
    {"name": "c", "geometry": Point(1, 1)},
], crs=4326)

# First two points:
points2 = points.iloc[:2, :]
buffers = points.buffer(0.1)
# GeoDataFrame of buffers with values of 1 and 2:
buffers = gpd.GeoDataFrame({
    "value": [1, 2, 3],
    "geometry": buffers
})
buffers.sjoin(points2, how='left')
```

# Questions and next steps

The code above shows that R and Python are similar in many ways.
Differences include:

- R's package **sf** use of the S2 spherical geometry engine by default for lon/lat data, which can mean fewer lines of code for some operations (e.g. buffers)
- The R package **sf** returns measures with units, which can make conversions easier, but which can also be confusing for new users
- Python uses the 'zip:' syntax for virtual file system access, while R uses '/vsizip/'
- Python has native support for interactive plotting with the `.explore()` method, while R requires an additional package such as **tmap** or **mapview** to create interactive plots

The above code is also a good example of how to create a reproducible workflow for geographic data analysis.
We hope that it can also act a bit like a [Rosetta Stone](https://en.wikipedia.org/wiki/Rosetta_Stone) for those who are familiar with one language and want to learn the other.

It also raises some questions, which we leave unanswered for the community to consider and comment on:

- Which language is more concise?
  - While there are slightly fewer lines of R code in the examples above, there may be ways to improve the Python code
- Which language runs quicker?
    - It would be interesting to see benchmarks for the above code, perhaps in a future [geocompx](https://geocompx.org/) blog post or even a multi-language benchmarking package
- Which language is quicker to write code in (the answer likely depends on your prior experience and tastes)?

We welcome input on these questions and any other comments on the above code, the source code of which can be found at [github.com/geocompx/geocompx.org](https://github.com/geocompx/geocompx.org/blob/main/post/2023/rgdal-retirement/index.qmd).
If you would like to contribute, with ideas, comments, or additional questions, feel free to get in touch via the [geocompx](https://geocompx.org/) website, our [Discord server](https://discord.gg/PMztXYgNxp), in a [GitHub Discussion](https://github.com/orgs/geocompx/discussions), on [Mastodon](https://fosstodon.org/tags/geocompx) or anywhere else.

# Further reading

There is lots more to learn in this space.
For more information, the following resources are recommended:

- [Geocomputation with R](https://r.geocompx.org/) 
- [Geocomputation with Python](https://py.geocompx.org/)
- [Spatial Data Science with applications in R and Python](https://r-spatial.org/python/), which provides R and Python code side-by-side
- A great tutorial that simultaneously covers R and Python is [Tools and packages to query and process Sentinel-1 and Sentinel-2 data with R and Python](https://github.com/loreabad6/ogh23) by Lorena Abad.
